---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# ReadabilityCzech

<!-- badges: start -->
[![Codecov test coverage](https://codecov.io/gh/cinkova/ReadabilityCzech/branch/master/graph/badge.svg)](https://app.codecov.io/gh/cinkova/ReadabilityCzech?branch=master)
[![CRAN status](https://www.r-pkg.org/badges/version/ReadabilityCzech)](https://CRAN.R-project.org/package=ReadabilityCzech)
[![R-CMD-check](https://github.com/cinkova/ReadabilityCzech/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/cinkova/ReadabilityCzech/actions/workflows/R-CMD-check.yaml)
<!-- badges: end -->

ReadabilityCzech computes readability of texts written in Czech with language-specific adaptations of the classic readability formulas [*Flesch Reading Ease* and *Flesch-Kincaid Grade Level*](https://en.wikipedia.org/wiki/Flesch%E2%80%93Kincaid_readability_tests).

Both readability formulas use counts of syllables, word counts per sentence, and sentence counts. The syllable counts per word are computed by a dedicated script in this package. The word and sentence counts are obtained from an automatic text analysis implemented by the [`udpipe` package by J. Wijffels](https://cran.r-project.org/web/packages/udpipe/vignettes/udpipe-universe.html).  


## Installation

You can install the development version of ReadabilityCzech like so:

```{r}
devtools::install_github("cinkova/ReadabilityCzech")
```

## Example
### Preprocessing
You want to gauge the readability of a Czech text. This is how you prepare it:

- Make sure that it is UTF-8 encoded and in the plain text format. 
- Use an NLP tool - a syntactic parser. It will
    - tokenize the text (ie. recognize word and sentence boundaries)
    - tell you the part of speech and lemma of each word
    - display syntactic relations between words (not needed in this use case). 

There are many NLP tools, but you need one that works with [Universal Dependencies](universaldependencies.org). 
For Czech we recommend [UDPipe](https://lindat.mff.cuni.cz/services/udpipe/info.php) with the language model `czech-pdt-udXXXX`. 

You can use UDPipe through:

- [Web GUI](https://lindat.mff.cuni.cz/services/udpipe/)
    
- [REST API](https://lindat.mff.cuni.cz/services/udpipe/api-reference.php) 
    
- R package [`udpipe`](https://CRAN.R-project.org/package=udpipe)


Note that the `udpipe` R package uses [UDPipe1](https://ufal.mff.cuni.cz/udpipe/1), whereas the GUI and API point to the current version of [UDPipe2](https://ufal.mff.cuni.cz/udpipe/2). However, for the readability use case this hardly makes any difference. 

The output from the parser will be in the so-called [CONLL-U](https://universaldependencies.org/format.html) format, which looks like this:

```
# generator = UDPipe 2, https://lindat.mff.cuni.cz/services/udpipe
# udpipe_model = english-ewt-ud-2.12-230717
# udpipe_model_licence = CC BY-NC-SA
# newdoc
# newpar
# sent_id = 1
# text = It is ok.
1	It	it	PRON	PRP	Case=Nom|Gender=Neut|Number=Sing|Person=3|PronType=Prs	3	nsubj	_	TokenRange=0:2
2	is	be	AUX	VBZ	Mood=Ind|Number=Sing|Person=3|Tense=Pres|VerbForm=Fin	3	cop	_	TokenRange=3:5
3	ok	ok	ADJ	JJ	Degree=Pos	0	root	_	SpaceAfter=No|TokenRange=6:8
4	.	.	PUNCT	.	_	3	punct	_	TokenRange=8:9
```

```{r warning=FALSE, message=FALSE}
library(ReadabilityCzech)
library(udpipe)
library(tidyverse)
library(magrittr)

```

We are going to compute the readability of a passage from a freely available e-book in Czech:

```{r}
plaintext <- read_lines("https://www.gutenberg.org/cache/epub/29648/pg29648.txt") %>% 
  stringr::str_flatten(collapse = " ") %>% str_sub(start = 3008, end = 7508)
```

Here are the first 505 characters of `plaintext`. 
```{r}
plaintext %>% str_sub(start = 1, end = 505)
```

### Three ways to obtain a `.conllu` file from your plain text

This is one of the possible ways to obtain the conll-u file using the REST-API. If you choose this way, you will have to mount specialized libraries for API handling. 

```{r}
library(RCurl)
library(jsonlite)

proresult <- RCurl::postForm(uri = "https://lindat.mff.cuni.cz/services/udpipe/api/process", 
                             .params = c(data = plaintext, tokenizer = TRUE, 
                                         tagger = TRUE, parser = TRUE, 
                                         model = "czech-pdt-ud-2.12-230717") ) %>%
  fromJSON() 
result <- proresult$result
write_lines(result, file = "data-raw/result_plaintext.conllu")

```

Another way would be with the `udpipe` library. Here is a cookbook. For more details, refer to the `udpipe` package documentation. 

First download the appropriate language model (naming just the language picks the largest if several are available).
To see the list of models, you have to consult the [UDPipe1 documentation](https://ufal.mff.cuni.cz/udpipe/1). 

```{r message=FALSE}
czech_model <- udpipe::udpipe_download_model(language = "czech", model_dir = "data-raw")
```
Then load (i.e. activate) the model. 

```{r }
czech <- udpipe::udpipe_load_model(czech_model$file_model)
```

Eventually, run the parser locally on your computer, using the model you have downloaded and then loaded.

```{r}
plaintext_udpipeR <- udpipe::udpipe_annotate(object = czech, x = plaintext, 
                                        tokenizer = "tokenizer", 
                                        tagger = "default", 
                                        parser = "default")
```

You could save some time with setting `parser` to `"none"`, when you process large data. 

```{r}
result_udpipeR <- plaintext_udpipeR$conllu
write_lines(result_udpipeR, "data-raw/result_plaintext_udpipeR.conllu")
```

Since each of these methods uses a different version of the language model and tool, the files are not identical (but both do a good job). 

If you do not mind fiddling with the UDPipe web GUI, upload a plain text file or copy-paste a plain text in the GUI, select the Czech model, hit the Process Input button and then the Save Output File button. 

Whichever way you have used, at this point you should have a file with the suffix `.conllu`. 


### Converting the `.conllu` file to the conll-u data frame
For this this task, you need the `udpipe` package.  

```{r}
plaintext_df <- udpipe_read_conllu(file = "data-raw/result_plaintext.conllu")
glimpse(plaintext_df)
```


## Computing readability 

### Flesch Reading Ease

```{r}
compute_FRE(plaintext_df)
```

### Flesch-Kincaid Grade Level

```{r}
compute_FleschKincaidGradeLevel(plaintext_df)
```

## Other text statistics
Both readability scores are based on sentence, word, and syllable counts. If you are interested in these statistics, you can use functions that were built to support `compute_FRE()` and`compute_FleschKincaidGradeLevel()`.

```{r}
count_sentences_in_conllu(plaintext_df)
```

```{r}
count_average_sentence_length(plaintext_df)
```

```{r}
count_words_in_conllu(plaintext_df)
```

```{r}
count_average_token_length_in_syllables(plaintext_df)
```

```{r}
create_syllablecount_per_token_df(plaintext_df) %>% slice_head(n = 10)
```


## Acknowledgments
Supported by the Czech Science Foundation grant 19-19191S: Linguistic Factors of Readability in Czech Administrative and Educational Texts. The work described herein has also been using data/tools/services provided by the LINDAT/CLARIAH-CZ
Research Infrastructure (https://lindat.cz) (Project No. LM2018101) and Czech National Corpus (Project No. LM2023044), both supported by the Ministry of Education, Youth and Sports of the Czech Republic. 

If you use this package, please cite this paper:

```
@inproceedings{biblio:BeCiAdaptationClassic2021,
	address = {Cham, Switzerland},
	series = {Lecture Notes in Computer Science},
	title = {{Adaptation of Classic Readability Metrics to Czech}},
	volume = {12848},
	copyright = {All rights reserved},
	isbn = {978-3-030-83526-2},
	booktitle = {{Lecture Notes in Artificial Intelligence, 24th International Conference on Text, Speech and Dialogue}},
	publisher = {Springer},
	author = {Bendová, Klára and Cinková, Silvie},
	editor = {Ekštein, Kamil and Pártl, František and Konopík, Miroslav},
	year = {2021},
	note = {tex.organization: University of West Bohemia},
	pages = {159--171},
}

```
